\tcode{Generalisation of nth_element to a range of nths}

### 2021-12-25 draft_lundberg_nth_elements.tex
updated tex as explained on first page

\newenvironment{MyQuote}{%
    \begin{myQuoteEnumerate}[resume=*,series=MyQuoteSeries]%
    \color{blue}%
    \item \begin{quote}%
}{%
    \end{quote}%
    \end{myQuoteEnumerate}%
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{lineno}
%\iffalse
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{lipsum}
\makeatletter
\renewcommand{\linenumberfont}{\fontsize{12pt}{3pt}\normalfont\smaller\color{gray}}
\makeatother
%\usepackage{lineno}
%\linenumbers
%\fi
%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1.2}
%%%%%%%%%%%%%%%%%%%%%%%%
\makeatletter
\renewcommand\tableofcontents{%
    \@starttoc{toc}%
}
\makeatother

@@

\begin{tabular}{ll}
Document \#:& D2375R1\\
Date: &2021-12-19 \\
Project: & Programming Language C++ \\
Audience: & LEWG \\
Email: &\href{mailto:lundberj@gmail.com}{lundberj@gmail.com}
\end{tabular}

@@

\subsection*{Contents}
\tableofcontents

@@

Draft 2021-12-16: Added clarifications and more background, explanations, wording clarifications, references, summarized performance study, applications, more questions/answers. --Based on feedback and material sent to SG9 list 2021-12-09. Synopsis: \tcode{S last} argument was missing 2021-12-19.

@@

The tony table with this proposal, \emph{After}, for all examples are the same:\newline

\begin{tabular}{|l|} 
  \hline 
  \textbf{After} \\
  \hline 
\begin{codeblock} 
nth_elements(v, nths, pred);
\end{codeblock} 
\\
\hline 
\end{tabular}

@@

\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
nth_element(v.begin(), nths[1], v.end(), pred);
nth_element(v.begin(), nths[0], nths[1], pred);
nth_element(nths[1]+1, nths[2], v.end(), pred);
\end{codeblock} 
\\
\hline 
\end{tabular}

@@

\subsection*{Alternative 1b: Hand-wired for size 3. O(N $\cdot$ M)}

@@

\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
nth_element(v.begin(), nths[0],v.end(), pred);
nth_element(nths[0], nths[1], v.end(), pred);
nth_element(nths[1], nths[2], v.end(), pred);
\end{codeblock} 
\\
\hline 
\end{tabular}

@@

\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
sort(v,pred);
\end{codeblock}
\\
\hline 
\end{tabular}

@@

A small sorted windows into a large data set can be selected as if sorted by partitioning at two points. For example, if `j` items fit on a display page, we can jump to page \tcode{k}, that is, into the range from \tcode{a=v.begin()+j*k} to \tcode{b=v.begin()+j*(k+1)}:

@@

\enlargethispage*{-8em}

@@

\enlargethispage*{-5em}
\subsection{Histogram normalization and bin selection. Application to image equalization}

Image equalization, or data equalization in general is a data processing technique used to enhanced contrast.\footnote{
The subject described for example at \url{https://en.wikipedia.org/wiki/Histogram_equalization} and \url{https://www.tutorialspoint.com/dip/histogram_equalization.htm}}

Equalization makes use of the [Cumulative distribution function, CDF\footnote{CDF: \url{https://en.wikipedia.org/wiki/Cumulative_distribution_function}} of the image (or data in general). The cumulative distribution can be estimated by taking the cumulant of a \emph{histogram} of the data. Finding good bins is itself done by histogram normalization, which itself relies on an estimated CDF.

An interesting alternative to cumulant-of-histogram, is to calculate or estimate the CDF from the ordered data directly. If this is done with a full sort, and exact descriptive CDF is obtained, resulting in a loss-less equalization (modulo floating point and image format aspects).%
\footnote{A more detailed description of sort-based image equalization: Image enhancement using sorted histogram specification and POCS postprocessing -
Il-Lyong Jung and Chang-Su Kim, 2007 IEEE}.

As a performance optimization we can avoid a full sort, and instead use \tcode{nth_elements} and ensure exact results at specific cumulant values (partition points) in the CDF, such as at 0.25, 0.5, 0.75. That is, we use a range-of-nth partitioning to approximate the full sorted data. Worked through examples are found at \cite{p2375RefImpl}.

%Example image: Einstein. Original (left), equalized using sort-based CDF (right)
%
%
%Again, the original (left), compared to equalization using approximate CDF using **range-of-nths** partitioning with `m=7`, `m=5`, or `m=3` partition points.
%
%
%The exact CDF (using sort) is shown in gray, along with the `range-of-nths` approximations.
%
%Example image: Tennforsen (Sweden). Original (left), equalized using sort-based CDF (right)
%
%original, vs sort based:
%
%\begin{tabular}{|l|l|} 
%\hline
%Original: & Equalized using full sort\\
%\includegraphics[width=0.50\textwidth]{plotting/examples/forsen.png} &
%\includegraphics[width=0.50\textwidth]{plotting/examples/forsen_sort.png} \\
%\hline
%Equalized using range-of-nths: m=3: & Equalized using range-of-nths: m=3\\
%\includegraphics[width=0.50\textwidth]{plotting/examples/forsen_partition2.png} &
%\includegraphics[width=0.50\textwidth]{plotting/examples/forsen_partition5.png}\\
%\hline
%\end{tabular}
%
%Again, the original (left), compared to equalization using approximate CDF using **range-of-nths** partitioning with `m=5`, `m=3`, or `m=3` partition points:
%
%original, vs `m=5`:
%
%`m=3`, vs `m=2`:
%
%The exact CDF (using sort) is shown in gray, along with the `range-of-nths` approximations.
%
%\includegraphics[width=0.21\textwidth]{plotting/examples/forsen_cdf_approximation.png}
%

\section{Wording and Synopsis }

@@

@@

\textbf{2} Preconditions: [first, nth) and [nth, last) are valid ranges. For the overloads in namespace std, RandomAccessIterator meets the Cpp17ValueSwappable requirements ([swappable.requirements]), and the type of *first meets the Cpp17MoveConstructible (Table 30) and Cpp17MoveAssignable (Table 32) requirements.
\added{For the overloads taking a range [nths_first,nths_last), \newline RandomAccessIterator2 is a RandomAccess iterator, and *nths_first is convertible to \newline RandomAccessIterator. For every iterator i and j in the range [nths_first,nths_last), it holds that if (i<j) then !(*j<*i).}

@@

\textbf{4} Effects: After nth_­element the element in the position pointed to by nth is the element that would be in that position if the whole range were sorted with respect to comp and proj, unless nth==last. Also for every iterator i in the range [first, nth) and every iterator j in the range [nth, last) it holds that: bool(invoke(comp, invoke(proj, *j), invoke(proj, *i))) is false.

\textbf{5} 
Complexity: For the overloads with no ExecutionPolicy, linear on average.
For the overloads with an ExecutionPolicy, O(N) applications of the predicate, and O(N log N) swaps, where N = last - first. 
\added{For overloads taking a range [nths_first,nths_last) but no ExecutionPolicy the complexity is approximately \mbox{O(N log m + M)} where m is the number of unique elements in [nths_first,nths_last) where M = nths_last - nths_first.
For overloads taking a range [nths_first,nths_last) and an ExecutionPolicy the complexity is \mbox{O(N log N + M)}.}

@@

Added signatures:

@@

namespace ranges {
  template<random_access_iterator I, sentinel_for<I> S, 
  random_access_range R2, class Comp = ranges::less, class Proj = identity>
  requires sortable<I, Comp, Proj>
  constexpr I nth_element(I first, R2&& nths, S last, Comp comp = {}, Proj proj = {});

@@

\section{Questions and Answers}

@@

\subsection{Q: Is there's a need to require `nths` be `sized_range`?}

A: Not sure, I don't think so. The example implementation does not use .size(). I first had a different answer to this but updated after further comments.

\subsection{Q: How should the `nths` be provided?}

The current-standard single-nth version uses a single iterator `nth` to designate the location in the range. A range of iterators seems to me the most natural way to designate multiple arbitrary locations in a range.

@@

\subsection{Q: Request for more motivation. Benefit and performance beyond Ordo.}

>>>>

\newenvironment{MyQuote}{%
    \begin{myQuoteEnumerate}[resume=*,series=MyQuoteSeries]%
    \color{blue}%
    \item \begin{quote}%
}{%
    \end{quote}%
    \end{myQuoteEnumerate}%
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{lineno}
%\iffalse
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{lipsum}
\makeatletter
\renewcommand{\linenumberfont}{\fontsize{12pt}{3pt}\normalfont\smaller\color{gray}}
\makeatother
%\usepackage{lineno}
%\linenumbers
%\fi
%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1.0}
%%%%%%%%%%%%%%%%%%%%%%%%
\makeatletter
\renewcommand\tableofcontents{%
    \@starttoc{toc}%
}
\makeatother

@@

\begin{tabular}{ll}
Document \#:& D2375R1\\
Date: &2021-12-25 \\
Project: & Programming Language C++ \\
Audience: & LEWG \\
Email: &\href{mailto:lundberj@gmail.com}{lundberj@gmail.com}
\end{tabular}

@@

\subsection*{Contents}
\renewcommand{\baselinestretch}{1.0}\normalsize
\tableofcontents
\renewcommand{\baselinestretch}{1.2}\normalsize

@@

2021-12-25: Simplified wording (change marked at section \ref{synopsis}). Added images for equalization example. Formatting and typos. Added example of \emph{Before} with projections, and example \emph{Before} with fixed size m=5.\newline 
2021-12-16: Added clarifications and more background, explanations, wording clarifications, references, summarized performance study, applications, more questions/answers. --Based on feedback and material sent to SG9 list 2021-12-09. Synopsis: \tcode{S last} argument was missing 2021-12-19.

@@

For all examples, the Tony Tables--\emph{After} case is the same:\newline

\hspace{3ex}\begin{tabular}{|l|} 
  \hline 
  \textbf{$\star \star$ After $\star \star$} \\
  \hline 
\begin{codeblock} 
nth_elements(v, nths, pred);
\end{codeblock}\\[1em]
or, for the examples using a projection:
\\
\begin{codeblock} 
nth_elements(v, nths, pred, proj);
\end{codeblock} 
\\
\hline 
\end{tabular}

@@

\hspace{3ex}\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
nth_element(v.begin(), nths[1], v.end(), pred);
nth_element(v.begin(), nths[0], nths[1], pred);
nth_element(nths[1]+1, nths[2], v.end(), pred);
\end{codeblock} 
\\
\hline 
\end{tabular}

@@

If we are using a projection, we must use sub-ranges or something to the same effect (and with the same risk of corner-case bugs, eg missed +1 at the right places):

\hspace{3ex}\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
ranges::nth_element(v, nths[1], pred, proj);
ranges::nth_element(sub_range(v.begin(),nths[1]), nths[0], pred, proj);
ranges::nth_element(sub_range(nths[1]+1,v.end()), nths[2], pred, proj);
\end{codeblock} 
\\
\hline 
\end{tabular} 
\subsection*{Alternative 1b: Hand-wired bisection for nths of known size 5. O(N) but messy}

\hspace{3ex}\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
nth_element(v.begin(), nths[5/2-1], v.end(), pred);
nth_element(v.begin(), nths[(5/2)/2-1], nths[5/2-1], pred);
nth_element(nths[(5/2)/2-1]+1, nths[5/2-1], v.end(), pred);
nth_element(nths[(5/2)-1]+1, nths[(5/2)/2+5/2-1], nths[5-1]), pred);
nth_element(nths[5-1]+1, nths[(5/2)/2+5/2-1], v.end(), pred);
\end{codeblock} 
\\
\hline 
\end{tabular}

\subsection*{Alternative 1c: Hand-wired for size 3. O(N $\cdot$ M)}

@@

\hspace{3ex}\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
nth_element(v.begin(), nths[0],v.end(), pred);
nth_element(nths[0]+1, nths[1], v.end(), pred);
nth_element(nths[1]+1, nths[2], v.end(), pred);
\end{codeblock} 
\\
\hline 
\end{tabular}

@@

\hspace{3ex}\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
sort(v,pred);
\end{codeblock}
\\
\hline 
\end{tabular}

@@

A small sorted windows into a large data set can be selected as if sorted by partitioning at two points. For example, if \tcode{j} items fit on a display page, we can jump to page \tcode{k}, that is, into the range from \tcode{a=v.begin()+j*k} to \tcode{b=v.begin()+j*(k+1)}:

@@

@@

\newpage

\subsection{Histogram equalization and bin selection. Application to image equalization}

Image equalization, or data equalization in general is a data processing technique used to enhanced contrast.\footnote{
The subject is described for example at \url{https://en.wikipedia.org/wiki/Histogram_equalization} and \url{https://www.tutorialspoint.com/dip/histogram_equalization.htm}} 
Equalization makes use of the Cumulative distribution function, CDF\footnote{CDF: \url{https://en.wikipedia.org/wiki/Cumulative_distribution_function}} of the image (or data in general). The cumulative distribution can be estimated by taking the cumulant of a \emph{histogram} of the data. Finding good bins is itself done by histogram equalization, which itself relies on an estimated CDF.

An interesting alternative to cumulant-of-histogram, is to obtain the CDF from ordering the data directly. If this is done with a full sort, and exact descriptive CDF is obtained, resulting in a loss-less equalization (modulo floating point and image format aspects).%
\footnote{A more detailed description of sort-based image equalization: Image enhancement using sorted histogram specification and POCS postprocessing -
Il-Lyong Jung and Chang-Su Kim, 2007 IEEE International Conference on Image Processing, ICIP 2007 Proceedings.}

As a performance optimization we can avoid a full sort, and instead use \tcode{nth_elements} and ensure exact results at specific cumulant values (partition points) in the CDF, such as at 0.25, 0.5, 0.75. That is, we use a range-of-nth partitioning to approximate the full sorted data. Worked through examples are found at \cite{p2375RefImpl}. In summary, consider the following image of Tännforsen (Sweden's greatest waterfall), with three different equalization methods.

\begin{tabular}{|l|l|} 
\hline
Original: & Equalized using full sort\\
\includegraphics[width=0.50\textwidth]{plotting/examples/forsen_roundtrip.small.jpg} &
\includegraphics[width=0.50\textwidth]{plotting/examples/forsen_sort.small.jpg} \\
\hline
Equalized using range-of-nths: m=2: & Equalized using range-of-nths: m=5\\
\includegraphics[width=0.50\textwidth]{plotting/examples/forsen_partition2.small.jpg} &
\includegraphics[width=0.50\textwidth]{plotting/examples/forsen_partition5.small.jpg}\\
\hline
\end{tabular}
\enlargethispage*{1em}

\begin{center}
\includegraphics[width=0.55\textwidth]{plotting/examples/forsen_cdf_approximation.png}
\end{center}
Above, the exact CDF (using sort) of the original image is shown in gray, along with approximations using \emph{range-of-nths}. At m=2, the equalization algorithm is equivalent to mapping the image min,max to grayscale values zero and 1.

\newpage
\section{Wording and Synopsis }

@@

\added{Underlined green text} marks additions.

@@

\textbf{2} Preconditions: [first, nth) and [nth, last) are valid ranges. For the overloads in namespace std, RandomAccessIterator meets the Cpp17ValueSwappable requirements ([swappable.requirements]), and the type of *first meets the Cpp17MoveConstructible (Table 30) and Cpp17MoveAssignable (Table 32) requirements.
\added{For the overloads taking a range [nths_first,nths_last), \newline RandomAccessIterator2 is a RandomAccess iterator, and *nths_first is convertible to \newline RandomAccessIterator.}

@@

\textbf{4} Effects: After nth_­element the element in the position pointed to by nth is the element that would be in that position if the whole range were sorted with respect to comp and proj, unless nth==last. Also for every iterator i in the range [first, nth) and every iterator j in the range [nth, last) it holds that: bool(invoke(comp, invoke(proj, *j), invoke(proj, *i))) is false. 
\added{For the overloads taking a range of nths, this holds for all nth in nths}$^\ddagger$

\textbf{5} 
Complexity: For the overloads with no ExecutionPolicy, linear on average.
For the overloads with an ExecutionPolicy, O(N) applications of the predicate, and O(N log N) swaps, where N = last - first. 
\added{For overloads taking a range [nths_first,nths_last) but no ExecutionPolicy the complexity is approximately \mbox{O(N log m + M)} where m is the number of unique elements in [nths_first,nths_last), and M = nths_last - nths_first.
For overloads taking a range [nths_first,nths_last) and an ExecutionPolicy the complexity is \mbox{O(N log N + M)}.}

$\ddagger$: \textbf{Note:} Rev R0 had an different wording with the same intention, but was redundant, as well as incomplete with respect to \emph{comp} and \emph{proj}.

@@

Added signatures \tcode{std::}

@@

\end{codeblockAdd}

Added signatures \tcode{std::ranges::}

\begin{codeblockAdd}
namespace ranges {
  template<random_access_iterator I, sentinel_for<I> S, 
  random_access_range R2, class Comp = ranges::less, class Proj = identity>
  requires sortable<I, Comp, Proj>
  constexpr I nth_element(I first, R2&& nths, S last, Comp comp = {}, Proj proj = {});

@@

\newpage
\section{Questions and Answers}

@@

\subsection{Q: Is there's a need to require \tcode{nths} be \tcode{sized_range}?}

A: Not sure, I don't think so. The example implementation does not use .size().

\subsection{Q: How should the \tcode{nths} be provided?}

The current-standard single-nth version uses a single iterator \tcode{nth} to designate the location in the range. A range of iterators seems to me the most natural way to designate multiple arbitrary locations in a range.

@@

\subsection{Q: Benefits and performance beyond Ordo.}
